Lesson 5
========================================================
```{r load data}
library(ggplot2)
library(dplyr)

setwd("~/src/udacity-nano-da-04")

pf = read.csv("pseudo_facebook.tsv", sep="\t")
names(pf)
```
### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) +
  geom_boxplot() +
  stat_summary(fun.y = mean, geom = "point", shape = 4)

ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(gender))) +
  geom_line(aes(color = gender), stat = "summary", fun.y = median)

pf.fc_by_age_gender <- pf %>% 
  filter(!is.na(gender)) %>%
  group_by(age, gender) %>%
  summarise(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n = n()) %>%
  ungroup() %>%
  arrange(age)

# Same mean plot, only using summarized data.
ggplot(aes(x = age, y = mean_friend_count),
       data = pf.fc_by_age_gender) +
  geom_line(aes(color = gender))

```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}

```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}
install.packages("tidyr")
library(tidyr)

# Reshape from long to wide format ("male" and "female" become columns).
spread(
  subset(pf.fc_by_age_gender, 
         select = c('gender', 'age', 'median_friend_count')),
  gender,
  median_friend_count)

install.packages('reshape2')
library(reshape2)

pf.fc_by_age_gender.wide <- dcast(
  pf.fc_by_age_gender,
  age ~ gender,
  value.var = "median_friend_count"
)

head(pf.fc_by_age_gender.wide)

ggplot(pf.fc_by_age_gender.wide,
       aes(x = age, y = female / male)) +
  geom_line() +
  geom_hline(yintercept = 1, alpha = .3, linetype = 2)

```


***

### Ratio Plot
Notes:

```{r Ratio Plot}

```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
pf$year_joined <- floor(2014 - (pf$tenure / 365))

summary(pf$year_joined)
```

***

### Cut a Variable
Notes:
# Create a new variable in the data frame
# called year_joined.bucket by using
# the cut function on the variable year_joined.

# You need to create the following buckets for the
# new variable, year_joined.bucket

#        (2004, 2009]
#        (2009, 2011]
#        (2011, 2012]
#        (2012, 2014]

```{r Cut a Variable}

pf$year_joined.bucket <- cut(
  pf$year_joined, 
  breaks = c(2004, 2009, 2011, 2012, 2014))

table(pf$year_joined.bucket, useNA = "ifany")

```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}

ggplot(subset(pf, !is.na(year_joined.bucket)),
       aes(x = age, y = friend_count)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = median)

```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}
ggplot(subset(pf, !is.na(year_joined.bucket)),
       aes(x = age, y = friend_count)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = median) +
  geom_line(stat = "summary",
            fun.y = mean,
            linetype = 2)
  

```

***

### Friending Rate
Notes:

```{r Friending Rate}
with(
  subset(pf, tenure > 0),
  summary(friend_count / tenure))

```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}
ggplot(subset(pf, tenure > 0),
       aes(x = tenure, y = friendships_initiated / tenure)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = tenure, 
           y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_smooth(aes(color = year_joined.bucket))
```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

```{r Yogurt!}
yo <- read.csv("yogurt.csv")
yo$id <- factor(yo$id)
str(yo)

```

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
qplot(x = price, data = yo)
summary(yo)
length(unique(yo$price))
table(yo$price)
```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)

summary(yo$all.purchases)

```

***

### Prices over Time
Notes:

```{r Prices over Time}
ggplot(aes(x = time, y = price), data = yo) +
  geom_point(color = "blue",
             alpha = 1/20,
             position = position_jitter())
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(4230)
sample.ids <- sample(levels(yo$id), 16)
table(sample.ids)

ggplot(aes(x = time, y = price),
       data = subset(yo, id %in% sample.ids)) +
  facet_wrap(~id) +
  geom_line() +
  geom_point(aes(size = all.purchases), pch = 1)

set.seed(4000)
sample.ids <- sample(levels(yo$id), 16)
table(sample.ids)

ggplot(aes(x = time, y = price),
       data = subset(yo, id %in% sample.ids)) +
  facet_wrap(~id) +
  geom_line() +
  geom_point(aes(size = all.purchases), pch = 1)


```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:

```{r Matrix}
install.packages("GGally")
library(GGally)
theme_set(theme_minimal(20))

set.seed(1836)
pf_subset <- pf[ , c(2:7)]
names(pf_subset)

ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000), ],
        axisLabels = 'internal')

```

***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
str(nci)
colnames(nci) <- c(1:64)
```

```{r}
library(reshape2)

nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

