Lesson 3
========================================================

***

### What to Do First?
Notes:

getwd()
list.files()

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
pf <- read.csv("pseudo_facebook.tsv", sep="\t")
names(pf)
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
install.packages('ggplot2')
library(ggplot2)

qplot(x = dob_day, data = pf) + 
  scale_x_continuous(breaks=1:31) + 
  facet_wrap(~dob_month, ncol=3)

install.packages('ggthemes', dependencies = TRUE)
library(ggthemes)
theme_set(theme_minimal(24))
```

***

#### What are some things that you notice about this histogram?
Response:

***

### Moira's Investigation
Notes:

***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:

#### How many of your friends do you think saw that post?
Response:

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:

```{r Faceting}

```

#### Let’s take another look at our plot. What stands out to you here?
Response:

***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(x = friend_count, data = pf)
```

#### How is this plot similar to Moira's first plot?
Response:

***

### Limiting the Axes
Notes:

```{r Limiting the Axes}
qplot(x = friend_count, data = pf) +
  scale_x_continuous(limits = c(0, 1000))
```

### Exploring with Bin Width
Notes:

***

```{r Adjusting the Bin Width}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = c(0, 1000, 50)) +
  facet_wrap(~gender)
```

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50))
```

***

### Omitting NA Values
Notes:

```{r Omitting NA Values}
na.omit(pf)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Response:

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:

***

### Tenure
Notes:

```{r Tenure}
qplot(x = tenure, data = pf, color=I("black"), fill=I("#099DD9"),
      binwidth = 30)
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x = tenure/365, data = pf, 
      xlab = "Number of years using Facebook",
      ylab = "Number of users in sample",
      color=I("black"), fill=I("#F79420"),
      binwidth = .25) + 
  scale_x_continuous(breaks = seq(1, 7, 1), lim = c(0, 7))
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}

```

***

### User Ages
Notes:

```{r User Ages}
qplot(x = age, data = pf, 
      xlab = "Age",
      ylab = "Number of users in sample",
      color=I("black"), fill=I("#5760AB"),
      binwidth = 1) +
  scale_x_continuous(breaks = seq(0, 113, 10))
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
```{r}
qplot(x = friend_count, data = pf)

summary(pf$friend_count)

summary(log10(pf$friend_count + 1))

summary(sqrt(pf$friend_count))

install.packages('gridExtra')
library(gridExtra)

p1 <- ggplot(aes(x = friend_count), data = pf) + geom_histogram()

p2 <- p1 + scale_y_log10()

p3 <- p1 + scale_y_sqrt()

grid.arrange(p1, p2, p3, ncol = 1)
```

Notes:

***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}

```

***


### Frequency Polygons

```{r Frequency Polygons}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), 
      y = ..count../sum(..count..),
      xlab = "Friends count",
      ylab = "Proportion of users with that friend count",
      binwidth = 10, geom = 'freqpoly', color = gender) +
      scale_x_continuous(lim = c(0, 100), breaks = seq(0, 1000, 50))

ggplot(aes(x = friend_count, y = ..count../sum(..count..)), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend Count') + 
  ylab('Percentage of users with that friend count')
```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
qplot(x = www_likes, data = subset(pf, !is.na(gender)), 
      y = ..count../sum(..count..),
      xlab = "Likes",
      ylab = "Proportion of users with likes",
      geom = 'freqpoly', color = gender) +
      scale_x_continuous() +
      scale_x_log10()

ggplot(aes(x = www_likes), data = subset(pf, !is.na(gender))) + 
  geom_freqpoly(aes(color = gender)) + 
  scale_x_log10()
```

```{r Likes solution}
by(pf$www_likes, pf$gender, sum)
```

***

### Box Plots
Notes:

```{r Box Plots}
qplot(x = gender,
      y = friend_count,
      data = subset(pf, !is.na(gender)), 
      xlab = "Gender",
      ylab = "Friend count",
      geom = 'boxplot')
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}
qplot(x = gender,
      y = friend_count,
      data = subset(pf, !is.na(gender)), 
      xlab = "Gender",
      ylab = "Friend count",
      geom = 'boxplot',
      ylim = c(0, 1000))

qplot(x = gender,
      y = friend_count,
      data = subset(pf, !is.na(gender)), 
      xlab = "Gender",
      ylab = "Friend count",
      geom = 'boxplot') +
  scale_y_continuous(limits = c (0, 1000))

qplot(x = gender,
      y = friend_count,
      data = subset(pf, !is.na(gender)), 
      xlab = "Gender",
      ylab = "Friend count",
      geom = 'boxplot') +
  coord_cartesian(ylim = c (0, 250))

by(pf$friend_count, pf$gender, summary)
```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}
```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
qplot(x = gender,
      y = friendships_initiated,
      data = subset(pf, !is.na(gender)), 
      xlab = "Gender",
      ylab = "Friend requests",
      geom = 'boxplot') +
  coord_cartesian(ylim = c (0, 150))

by(pf$friendships_initiated, pf$gender, summary)
```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes > 0)

pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
```

Response:

***

### Analyzing One Variable

```{r}Diamonds
data(diamonds)

qplot(price, data = diamonds, binwidth = 1000)

dim(subset(diamonds, price >= 15000))

qplot(price,
      data = diamonds,
      binwidth = 500,
      xlim = c(0, 5000))

qplot(price,
      data = diamonds,
      binwidth = 100) +
      facet_wrap(~cut)

by(diamonds$price, diamonds$cut, max)
by(diamonds$price, diamonds$cut, min)
by(diamonds$price, diamonds$cut, median)

qplot(x = price, data = diamonds) +
  facet_wrap(~cut, scales = "free_y")

qplot(x = price / carat,
      data = diamonds,
      binwidth = 0.01) +
  scale_x_log10() +
  facet_wrap(~cut, scales = "free_y")
```

```{r}Diamonds boxplots by clarity

qplot(x = clarity,
      y = price,
      data = diamonds, 
      xlab = "Clarity",
      ylab = "Price",
      geom = 'boxplot') +
    coord_cartesian(ylim = c(600, 6100))
    
ggsave('diamonds-boxplot-clarity.png')

summary(subset(diamonds, color == 'D')$price)
summary(subset(diamonds, color == 'J')$price)

summary(diamonds$color)

IQR(subset(diamonds, color == 'D')$price)
IQR(subset(diamonds, color == 'J')$price)

```

Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!